---
title: "02. Simulating data by using MSOAC parameters"
author: "Quentin Pilard"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{02. Simulating data by using MSOAC parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
For applying the models, simulated data from the MSOAC was used.
The following code simulates recurrent event data for **relapses in RRMS and SPMS patients**. The data is then combined with baseline characteristics for RRMS and SPMS patients. The simulated data is saved as a CSV file for further use.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(purrr)
library(tibble)
library(reda)
library(tidyr)
library(gtsummary)
library(writexl)
library(kableExtra)
```

```{r message=FALSE, warning=FALSE}
# Set seed for reproducibility
set.seed(123)
```

```{r message=FALSE, warning=FALSE}
# Define a function to simulate recurrent event data using reda
simulate_recurrent_event_data <- function(n, max_event, mean_follow_up_time, max_follow_up_time, mstype, outcome) {
  
  simulate_individual_data <- function(id) {
    
    # Generate baseline hazard times (exponential distribution)
    basehaz <- rexp(n = pmax(pmin(round(rnorm(1, max_event/2, max_event/4)), max_event), 1), 
                    rate = 1 / mean_follow_up_time)
    basehaz <- basehaz[basehaz < max_follow_up_time]
    basehaz <- c(basehaz,max_follow_up_time)
    
    # Create a tibble for the individual's data
    tibble(
      id = id,
      status = c(rep(1, length(basehaz) - 1), 0),
      time = basehaz,
      mstype = mstype,
      outcome = outcome
    )
  }
  
  # Use purrr::map_dfr to apply the helper function to each individual and combine the results
  sim_data_all <- map_dfr(1:n, simulate_individual_data) %>%
    arrange(id, time)
  
  # Return the simulated data
  return(sim_data_all)
}
```

```{r message=FALSE, warning=FALSE}
# Generate the datasets for relapses in RRMS and SPMS patients
# this parameters have been changed in order to find differences between groups
# just for illustrative purposes
rrms_relapse_data<-simulate_recurrent_event_data(n = 791,
                                                 max_event = 10,
                                                 mean_follow_up_time = 8,
                                                 max_follow_up_time = 10, 
                                                 mstype = "RRMS", 
                                                 outcome = "RELAPSE")
spms_relapse_data<-simulate_recurrent_event_data(n = 522,
                                                 max_event = 4,
                                                 mean_follow_up_time = 5,
                                                 max_follow_up_time = 10, 
                                                 mstype = "SPMS", 
                                                 outcome = "RELAPSE")
spms_relapse_data$id <- spms_relapse_data$id + 791
```

```{r message=FALSE, warning=FALSE}
# Bind datasets
df <- bind_rows(rrms_relapse_data,
                spms_relapse_data)
```

```{r message=FALSE, warning=FALSE}
#Add baaseline characteristics
# Function to generate data for RRMS
generate_rrms_data <- function(n) {
  tibble(
    id = 1:n,
    Age = rnorm(n, mean = 36.5, sd = 9.1),
    Sex = factor(rbinom(n, 1, prob = 0.70), labels = c("Male", "Female")),
    Race = factor(rbinom(n, 1, prob = 0.88), labels = c("No white", "White")),
    Time_since_diagnosis = rlnorm(n, meanlog = log(2), sdlog = log(5/2)),
    EDSS_overall = rlnorm(n, meanlog = log(2), sdlog = log(3.5/1.5))
  )
}

# Function to generate data for SPMS
generate_spms_data <- function(n) {
  tibble(
    id = 1:n,
    Age = rnorm(n, mean = 49.4, sd = 8.1),
    Sex = factor(rbinom(n, 1, prob = 0.63), labels = c("Male", "Female")),
    Race = factor(rbinom(n, 1, prob = 0.96), labels = c("No white", "White")),
    Time_since_diagnosis = rlnorm(n, meanlog = log(14.5), sdlog = log(22/7.8)), # approximate log-normal
    EDSS_overall = rlnorm(n, meanlog = log(6), sdlog = log(6.5/4.5)) # approximate log-normal
  )
}
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Generate the datasets
rrms_data <- generate_rrms_data(791)
spms_data <- generate_spms_data(522)
spms_data$id <- spms_data$id + 791

#Merge with the previous part
df_revents <- inner_join(bind_rows(rrms_data, spms_data),df, by="id")
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
kbl(df_revents[1:10,]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

It is pertinent to mention that this type of data is not suitable for running recurrent events models (see [database layout section](C:/Users/david/revents/docs/articles/database_layout.html)). However, this raw data can be used for the first steps of data exploration as it is shown in the following section: [descriptive analyses](C:/Users/david/revents/docs/articles/descriptive_analyses.html).
